Visualizing distributions
========================================================

#### for Nicholas Reich's Statistical Modeling and Data Visualization course at UMass (PUBHLTH 690NR)

Being able to quickly and deftly charaterize univariate and bivariate distributions in a dataset is a crucial step towards understanding any dataset. 

We begin by loading the ggplot2 package, a cutting-edge, actively-developed package that helps create pretty graphics, and the MASS package, which has lots of datasets.
```{r}
library(ggplot2)
library(MASS)
```

Next, we're going to explore a dataset that is pre-loaded in R, the crabs dataset.
```{r}
data() ## look at all the datasets available to you
data(crabs) ## load the crabs dataset
str(crabs)
summary(crabs)
```

Notice that the 'sp' and 'sex' columns in the crabs dataframe are "Factors". A factor is the name of the data type that R uses for categorical variables. Before we dive into analyzing the data, let's learn a little bit about factors.
```{r factors}
table(crabs$sex)
head(crabs$sex)
levels(crabs$sex)
nlevels(crabs$sex)
str(crabs)
crabs$sex <- relevel(crabs$sex, ref="M") ## make 'Male' the reference level
str(crabs) # note the numeric values of $sex have changed
```
In general, R will read character columns in an imported dataset in as factors. However, on occasion, you will need to create a factor from scratch.  here is such an example:
```{r defineFactors}
f <- c("a", "b", "a", "b", "a", "c") ## here is a character vector
f.fac <- as.factor(f) ## we can make it a factor
as.numeric(f.fac) ## and then if we convert it to numeric you can see the underlying values
```

Here are some univariate explorations of the crabs data.
```{r univariateCrabs, message=FALSE}
qplot(CW, data=crabs)
qplot(CW, data=crabs, geom="histogram")

qplot(sp, data=crabs)
qplot(sp, data=crabs, fill=sex)
with(crabs, table(sp, sex))
```

Here is some exploration of continuous by continuous data.
```{r bivariateCrabs, message=FALSE}
qplot(CW, RW, data=crabs)
qplot(CW, RW, data=crabs, geom="point")
qplot(CW, RW, data=crabs, geom=c("point", "density2d"))

qplot(CW, RW, data=crabs, color=sex)
qplot(CW, RW, data=crabs, color=sp)

qplot(CW, FL, data=crabs, color=sex)
qplot(CW, FL, data=crabs, color=sp)
qplot(CW, FL, data=crabs, color=sp:sex)
qplot(CW, FL, data=crabs, facets= . ~ sp)
qplot(CW, FL, data=crabs, facets= . ~ sp, color=sex)


qplot(CW, RW, data=crabs)
qplot(CW, RW, data=crabs, geom=c("point"))
qplot(CW, RW, data=crabs, geom=c("point", "smooth"))
qplot(CW, RW, data=crabs, geom=c("point", "smooth"), color=sex)
qplot(CW, RW, data=crabs, geom=c("point", "smooth", "density2d"), color=sex)
```

You can also create interactive graphics using the manipulate() function. Note: these interactive graphs will not show up in the printout of code and graphs.

```{r manipulate, eval=FALSE}
library(manipulate)
manipulate(qplot(CW, crabs[,x], data=crabs), x=picker("RW", "FL", "CL", "CW", "BD"))
manipulate(qplot(CW, crabs[,x], data=crabs, color=sex), 
           x=picker("RW", "FL", "CL", "BD"))
manipulate(qplot(CW, crabs[,x], data=crabs, color=crabs[,y]), 
           x=picker("RW", "FL", "CL", "BD"),
           y=picker("sex", "sp"))
manipulate(qplot(CW, crabs[,x], data=crabs, color=crabs[,y])+xlim(15, xmax), 
           x=picker("RW", "FL", "CL", "BD"),
           y=picker("sex", "sp"),
           xmax=slider(25, 75))
```


And some categorical by continuous
```{r crabsFigs, message=FALSE}
qplot(sex, CW, data=crabs)
qplot(sex, CW, data=crabs, geom="point")
qplot(sex, CW, data=crabs, geom="boxplot")
qplot(sex, CW, data=crabs, geom="boxplot", facets=.~sp)
qplot(sex, CW, data=crabs, geom="boxplot", facets=sp~.)
qplot(sex, CW, data=crabs, geom="violin")
qplot(sex, CW, data=crabs, geom=c("violin", "point"))
```


```{r bigMatrix, message=FALSE}
require(GGally)
head(crabs)
crabs.mat <- crabs[,4:8]
ggpairs(crabs.mat)

```


```{r}
sessionInfo()
```
